### 2024-04-29 paper.tex
minor adjustments

\documentclass[twocolumn,10pt]{article}

\usepackage[a4paper,hmargin=1.5cm,vmargin=2cm]{geometry}
\setlength{\columnsep}{0.7cm}
%\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{balance}
\usepackage{palatino}
\usepackage[varqu]{inconsolata}

@@

\newcommand{\todo}[1]{[\textcolor{orange}{#1}]}
\newcommand{\new}[1]{#1} %\textcolor{blue}{#1}}

@@

\twocolumn[{%
\centering\phantomsection\addcontentsline{toc}{section}{Title}
\textbf{\Large Analyzing single-cell bisulfite sequencing data with \textit{MethSCAn}}\\[1.5ex]

Lukas P.~M. Kremer\textsuperscript{1,2,*},
Martina M. Braun\textsuperscript{2},
Svetlana Ovchinnikova\textsuperscript{1},
Leonie Küchenhoff\textsuperscript{1,2},\\
Santiago Cerrizuela\textsuperscript{2},
Ana Martin-Villalba\textsuperscript{2,*},
Simon Anders\textsuperscript{1,*}\\[1ex]
{\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex] 
\textsuperscript{2} Division of Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany}
\vspace{1.5ex}

version 2 /\!/ January 2024
\vspace{4.5ex}

\begin{minipage}{0.7\textwidth}
\small \setlength{\parindent}{1em}
\phantomsection\addcontentsline{toc}{subsection}{Abstract}
\noindent \textit{Abstract} -- Single-cell bisulfite sequencing (scBS) is a technique that enables the assessment of DNA methylation at single-base pair and single-cell resolution.
The analysis of large datasets obtained from scBS requires preprocessing to reduce data size, improve signal-to-noise ratio, and provide interpretability.
Typically, this is achieved by dividing the genome into large tiles and averaging the methylation signals within each tile.

@@

To facilitate the analysis of scBS data, we have developed a software tool called \textit{MethSCAn} that implements these methods and provides additional functionality.
\end{minipage}
\vspace{6.5ex}
}]
\setcounter{secnumdepth}{0}

\section{Introduction}

@@

%\phantomsection  % idk why we need this to silence hyperref's warning
\subsection{Standard approach}

@@

\section{Improved preprocessing of scBS data}

\subsection{Read-position aware quantitation} \label{residuals}

@@

\subsection{Finding variably methylated regions}

@@

\subsection{Application and benchmarks}

@@

%\phantomsection
\subsubsection{Correlating VMR methylation with gene expression}

@@

\section{Finding differentially methylated regions (DMRs)}

@@

%\phantomsection
\subsection{Detecting DMRs between oligodendrocytes and neural stem cells}

@@

\section{The \textit{MethSCAn} software toolkit}

@@

\section{Discussion and Conclusion}

@@

\subsection{Data Availability}

@@

\subsection{Code Availability}

@@

%\phantomsection
\section{Methods}

\subsection{Raw data}

@@

\subsection{Data storage}

@@

\subsection{Smoothing}

@@

\subsection{Methylation for an interval}

@@

\subsection{Finding differentially methylated regions}

@@

\subsection{Calculating cell to cell distances}

@@

\subsection{PCA with iterative imputation}

@@

\subsection{Analysis of scBS datasets for benchmarks}
To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{methscan prepare} and then smoothed with \texttt{methscan smooth} using the default bandwidth of 1000~bp.
This data was then analysed multiple times with different combinations of analysis methods, namely four ways to divide the genome into intervals, two ways to quantify methylation in these intervals, and four approaches for dimensionality reductions.
The following four sets of genomic intervals were used:
VMRs, obtained with \texttt{methscan scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$);
adjacent tiles of 100~kb width;
promoter regions as defined by the \textpm2~kb domain around the TSS of coding genes;
and candidate cis-regulatory regions annotated by the ENCODE consortium \citep{encode2020expanded}.
Methylation was quantified either by averaging binary methylation values, or by calculating the shrunken mean of the residuals as described earlier.

@@

\subsection{Mean neighbor score} \label{methods:score}
To assess performance of our methods, we employed a score that quantifies how well cell types (or cell states) are separated in 15-dimensional PCA space.
For data from \citet{luo2017single}, we used cell type labels that the authors manually curated based on CH-methylation.
For the multi-omic data set, we repeated the single-cell transcriptomics analysis described in \citet{kremer_scnmt} with two adjustments:
We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat} function \texttt{FindClusters(resolution = 0.1)}.
The score, based on the $\Gamma$-score \citep{Kireeva_2014}, varies between 0 and 1, where higher scores reflect better separation of cell types.
For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$, rounded.
The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.

\subsection{Correlation of DNA methylation and gene expression}
To assess correlation between gene expression and the methylation status of VMRs or promoters, we first detected VMRs with \texttt{methscan scan --bandwidth 1000 --var-threshold 0.05}.
We then quantified DNA methylation at VMRs and promoters with \texttt{methscan matrix}.
We defined promoters as $\pm2000$~bp regions centered on a genes transcription start site (TSS).
When multiple TSSs were annotated, we chose the TSS of the "principal" isoform \citep{appris}.
Log-normalized expression values reported by \citet{kremer_scnmt} were then correlated with methylation of the gene's promoter, or with methylation of the VMR closest to the gene body.
When multiple VMRs intersected the gene body, we chose the VMR with the highest methylation variance.
As a measure of methylation we used the shrunken mean of the residuals.
We omitted lowly expressed genes (with scRNA-seq counts in $<5\%$ of cells) and promoters and VMRs with low scBS coverage (in $<10$ cells).

@@

\phantomsection\addcontentsline{toc}{section}{Back Matter}
\subsection{Acknowledgments}

@@

\onecolumn 
\newpage
\section{Extended Data Figures}
\vspace{1.5cm}

>>>>

\documentclass[10pt]{article}

\usepackage[a4paper,hmargin=1.5cm,vmargin=2cm]{geometry}
%\setlength{\columnsep}{0.7cm}
%\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{balance}
\usepackage[varqu]{inconsolata}

@@

@@

{\centering\phantomsection\addcontentsline{toc}{section}{Title}
\textbf{\Large Analyzing single-cell bisulfite sequencing data with \textit{MethSCAn}}\\[1.5ex]

Lukas P.~M. Kremer\textsuperscript{1,2,*},
Martina M. Braun\textsuperscript{2},
Svetlana Ovchinnikova\textsuperscript{1},
Leonie Küchenhoff\textsuperscript{1,2},\\
Santiago Cerrizuela\textsuperscript{2},
Ana Martin-Villalba\textsuperscript{2,*},
Simon Anders\textsuperscript{1,*}\\[1ex]
{\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex] 
\textsuperscript{2} Division of Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany}
}
\vspace{4.5ex}

\section*{Abstract}

\noindent Single-cell bisulfite sequencing (scBS) is a technique that enables the assessment of DNA methylation at single-base pair and single-cell resolution.
The analysis of large datasets obtained from scBS requires preprocessing to reduce data size, improve signal-to-noise ratio, and provide interpretability.
Typically, this is achieved by dividing the genome into large tiles and averaging the methylation signals within each tile.

@@

To facilitate the analysis of scBS data, we have developed a software tool called \textit{MethSCAn} that implements these methods and provides additional functionality.

\section*{Introduction}

@@

\subsection*{Standard approach}

@@

\section*{Results}

\subsection*{Read-position aware quantitation} \label{residuals}

@@

\subsection*{Finding variably methylated regions}

@@

\subsection*{Application and benchmarks}

@@

\subsubsection{Correlating VMR methylation with gene expression}

@@

\section*{Finding differentially methylated regions (DMRs)}

@@

\subsection*{Detecting DMRs between oligodendrocytes and neural stem cells}

@@

\section*{The \textit{MethSCAn} software toolkit}

@@

\section*{Discussion}

@@

\subsection*{Data Availability}

@@

\subsection*{Code Availability}

@@

%\phantomsection
\section*{Methods}

\subsection*{Raw data}

@@

\subsection*{Data storage}

@@

\subsection*{Smoothing}

@@

\subsection*{Methylation for an interval}

@@

\subsection*{Finding differentially methylated regions}

@@

\subsection*{Calculating cell to cell distances}

@@

\subsection*{PCA with iterative imputation}

@@

\subsection*{Analysis of scBS datasets for benchmarks}
To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{methscan prepare} and then smoothed with \texttt{methscan smooth} using the default bandwidth of 1000~bp.
This data was then analysed multiple times with different combinations of analysis methods, namely four ways to divide the genome into intervals, two ways to quantify methylation in these intervals, and four approaches for dimensionality reductions.
The following four sets of genomic intervals were used:
VMRs, obtained with \texttt{methscan scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$);
adjacent tiles of 100~kb width;
promoter regions as defined by the \textpm2~kb domain around the TSS of coding genes;
and candidate cis-regulatory regions annotated by the ENCODE consortium \citep{encode2020expanded}.
Methylation was quantified either by averaging binary methylation values, or by calculating the shrunken mean of the residuals as described earlier.

@@

\subsection*{Mean neighbor score} \label{methods:score}
To assess performance of our methods, we employed a score that quantifies how well cell types (or cell states) are separated in 15-dimensional PCA space.
For data from \citet{luo2017single}, we used cell type labels that the authors manually curated based on CH-methylation.
For the multi-omic data set, we repeated the single-cell transcriptomics analysis described in \citet{kremer_scnmt} with two adjustments:
We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat} function \texttt{FindClusters(resolution = 0.1)}.
The score, based on the $\Gamma$-score \citep{Kireeva_2014}, varies between 0 and 1, where higher scores reflect better separation of cell types.
For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$, rounded.
The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.

\subsection*{Correlation of DNA methylation and gene expression}
To assess correlation between gene expression and the methylation status of VMRs or promoters, we first detected VMRs with \texttt{methscan scan --bandwidth 1000 --var-threshold 0.05}.
We then quantified DNA methylation at VMRs and promoters with \texttt{methscan matrix}.
We defined promoters as $\pm2000$~bp regions centered on a genes transcription start site (TSS).
When multiple TSSs were annotated, we chose the TSS of the "principal" isoform \citep{appris}.
Log-normalized expression values reported by \citet{kremer_scnmt} were then correlated with methylation of the gene's promoter, or with methylation of the VMR closest to the gene body.
When multiple VMRs intersected the gene body, we chose the VMR with the highest methylation variance.
As a measure of methylation we used the shrunken mean of the residuals.
We omitted lowly expressed genes (with scRNA-seq counts in $<5\%$ of cells) and promoters and VMRs with low scBS coverage (in $<10$ cells).

@@

\phantomsection\addcontentsline{toc}{section}{Back Matter}
\subsection*{Acknowledgments}

@@

\onecolumn 
\newpage
\section*{Extended Data Figures}
\vspace{1.5cm}
